If the Null Fits, You Must Omit: Measurement Artifacts Undermine Group Polarization Replications

### 2025-10-25 GroupPolarizationStatmod/paper/intro-latest.tex
finished corpus table

A relatively simple step could fix the problem: re-center the post-discussion
scale on each group’s own pre-discussion mean, aligning measurement with the
group’s frame rather than an arbitrary external ``neutral.'' This adjustment
restores the basic condition for inference: measuring change within a
consistent frame of reference.

But even if this measurement problem were fixed,
there are several problems in this kind of social science: Unstable
opinions, unmodeled input variance, and incommensurate experimental designs
all uniquely undermine any cumulative insight group polarization might
claim—problems known about and ignored for decades~\cite{Cartwright1973}. No
degree of scientific lawfulness can be established without evidence gathered
from rigorously modeled, commensurate experiments~\cite{Cartwright1999}.

@@

@@

In most group polarization experiments, opinions are measured on an
ordinal scale: Burnstein and Vinokur (1975) for example used a ten-point
scale from 1 to 10 representing “the lowest odds of success acceptable” to
try a certain behavior, where 1 means 10\%—Moscovici and Zavalloni (1969) gave participants a seven-point scale, from “strongly disagree” (-3) to “strongly agree” (+3) to report their opinions. In these and every other survey-based group polarization study I know, the reported opinions on the ordinal scale—with discrete numbered bins—were treated as if they were continuous-valued, or in other words as if they were the internal mental opinion value itself, whatever that might be neurobiologically. 
Liddell and Kruschke (2018) showed that in the general case this setup can
lead to false positives because ordinal scales cannot track changes in
extreme opinions beyond the ceiling of the measurement scale—this approach is
ubiquitous in social psychology and survey research, not at all limited to
group polarization.

We selected ten journal articles with sixty total experimental conditions
across the ten.

\begin{table}[ht]
\caption{\textbf{Journal articles in our corpus.}}
\centering
\begin{tabular}{clcc}
\toprule
Authors (Year) & Title & Journal & Category \\
\midrule
  Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes &
    Journal of Personality and Social Psychology & Politics \\
    Schkade, Sunstein, \& Hastie & When Deliberation Produces Extremism & Critical Review
                                 & Politics \\
Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts
In Individual Choice & Journal of Experimental Social Psychology & Persuasive Arguments \\

Burnstein \& Vinokur (1975) & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments explanation of choice shifts & Journal of Experimental Social Psychology & Persuasive Arguments \\
\bottomrule
\end{tabular}
\label{table:mr}
\end{table}

>>>>

A relatively simple step could fix the problem, we believe: measure twice
before discussion. Before forming groups, experimenters could poll opinions
using a scale centered arbitrarily at 0, with varying strengths of disagree
or agree with negative and positive integers, respectively. 
Then after likeminded groups are
selected, measure group opinions again before they discuss, but now with the
scale centered on the average group opinion. To keep the same number of bins,
add more bins to express more extreme views and take away bins from
the opposite side as necessary.

Even if this measurement problem were fixed, equally severe problems remain
for group polarization research. First there seems to be no stable 
``opinions'' per
se—it is more accurate to say people construct opinions when they are
prompted or need them for some other reason, as revealed by respondents
answering questions differently if the framing or ordering of them changes,
and by the fact that people respond differently to the same question over
time. These are just a few of several potential sources of variance that were never accounted for, which is also known to generate false detections of change. 
Finally, because group polarization experimental design was never
standardized its research corpus cannot cumulatively or systematically tell 
us anything general about the world—Dorian
Cartrwright (1973) tried to alert group polarization researchers to this
problem half a century ago~\cite{Cartwright1973}. General scientific claims
of the existence and functioning of some scientifically induced phenomenon
only work if the evidentiary observations were collected from
\emph{commensurate} experiments, or in other words, only from experiments
that share the exact same causal structure as expressed in a formal or 
mechanistic causal model~\cite{Cartwright1999}.

@@

We selected ten representative journal articles with sixty total experimental
conditions among them. For a representative sample of group polarization
research across decades, we selected two articles for each of five article
categories: \emph{Social Comparisons}, \emph{Persuasive Arguments},
\emph{Self-Categorization}, \emph{Social Decisions}, and \emph{Politics}. The
first four of these categories reference the theoretical tradition that the
authors were trying to prove right—the fifth, \emph{Politics}, was the
subject matter of the studies, but the two studies took no clear side in the
group polarization theory wars. This resulted in a representative sample of
group polarization research over time, as well: one article is from 1969,
four are from the 1970s, . Information about our corpus is summarized below
(Table~\ref{tab:articles}).  Please see the Supplemental Information

% \begin{table}[ht]

%   \caption{\textbf{Journal articles in our corpus.}}
%   \label{tab:articles}

%   \centering 
%   \rowcolors{2}{white}{gray!7}
%   \begin{tabular}{m{1.35in} m{2.45in}
%     m{1.3in} m{0.85in}}

%     \toprule

%     \textbf{Authors (Year)}& \textbf{Title} & \textbf{Journal} &
%     \textbf{Category} \\

%     \midrule 
%     Moscovici \& Zavalloni (1969) & The Group as a Polarizer of
%     Attitudes & \emph{JPSP} & Politics \\ 
%     Schkade, Sunstein, \& Hastie (2010) & When
%     Deliberation Produces Extremism & \emph{Critical~Review} & Politics \\

%     Bishop \& Myers (1970) & Discussion Effects on Racial Attitudes &
%     \emph{Science} & Social Comp's \\ 
%     Myers (1975) & Discussion-Induced Attitude Polarization & 
%     \emph{Human Relations} & Social Comp's \\

%     Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group
%     Induced Shifts In Individual Choice & \emph{JESP} & Persuasive Arg's \\
%     Burnstein \& Vinokur (1975) & What a person thinks upon learning he has
%     chosen differently from others: Nice evidence for the
%     persuasive-arguments\ldots & \emph{JESP} & Persuasive Arg's \\

%     Abrams, et al., (1990) & Knowing What to Think by Knowing Who You Are:
%     Self‐Categorization and the Nature of Norm Formation\ldots & 
%     \emph{Brit. J. of Soc. Psych.} & Self-Categ'n \\
%     Hogg, Turner, \& Davidson (1990) & Polarized Norms and Social Frames of 
%     Reference: A Test of the Self-Categorization Theory of Group 
%     Polarization & \emph{B. \& Appl. Soc. Psych.} & Self-Categ'n \\
%     Krizan \& Baron (2007) & Group Polarization and Choice Dilemmas: How
%     Important is Self-Categorization? & \emph{Eur. J. of Soc. Psych.} &
%     Self-Categ'n \\

%     Friedkin (1999) & Choice Shifts \& Group Polarization & 
%     \emph{Amer. Soc.  Rev.} & Soc'l Decisions \\

% \bottomrule
% \end{tabular}
% \end{table}
\begin{table}[ht]
\centering
\caption{\textbf{Journal articles in our corpus.}}
\label{tab:articles}

\begin{tabular}{m{0.9in} m{1.35in} m{2.45in} m{1.3in}}
\toprule
\textbf{Category} & \textbf{Authors (Year)} & \textbf{Title} & \textbf{Journal} \\
\midrule

\rowcolor{gray!10}
\multirowcell{2}{Politics} 
  & Moscovici \& Zavalloni (1969) & The Group as a Polarizer of Attitudes & \emph{JPSP} \\
\rowcolor{gray!10}
  & Schkade, Sunstein, \& Hastie (2010) & When Deliberation Produces Extremism & \emph{Critical~Review} \\[0.2em]

\rowcolor{gray!5}
\multirowcell{2}{Social Comp's \vspace{0.7em}}
% \multirowcell{2}{Social Comp's}
  & Bishop \& Myers (1970) & Discussion Effects on Racial Attitudes & \emph{Science} \\[0.2em]
\rowcolor{gray!5}
  & Myers (1975) & Discussion-Induced Attitude Polarization & \emph{Human Relations} \\[0.3em]

\rowcolor{gray!10}
\multirowcell{2}{Persuasive Arg's}
  & Burnstein \& Vinokur (1973) & Testing Two Classes of Theories About Group Induced Shifts In Individual Choice & \emph{JESP} \\
\rowcolor{gray!10}
  & Burnstein \& Vinokur (1975) & What a person thinks upon learning he has chosen differently from others: Nice evidence for the persuasive-arguments\ldots & \emph{JESP} \\

\rowcolor{gray!5}
\multirowcell{3}{Self-Categ'n}
  & Abrams, et al., (1990) & Knowing What to Think by Knowing Who You Are: Self‐Categorization and the Nature of Norm Formation\ldots & \emph{Brit. J. of Soc. Psy.} \\
\rowcolor{gray!5}
  & Hogg, Turner, \& Davidson (1990) & Polarized Norms and Social Frames of Reference: A Test of the Self-Categorization Theory of Group Polarization & \emph{B. \& Appl. Soc. Psy.} \\
\rowcolor{gray!5}
  & Krizan \& Baron (2007) & Group Polarization and Choice Dilemmas: How Important is Self-Categorization? & \emph{Eur. J. of Soc. Psy.} \\

\rowcolor{gray!10}
\multirowcell{1}{Soc'l Decisions}
  & Friedkin (1999) & Choice Shifts \& Group Polarization & \emph{Amer. Soc. Rev.} \\

\bottomrule
\end{tabular}
\end{table}

@@

In studies of group polarization, opinion measurements are typically ordinal, reflecting
ordered positions rather than measurable distances between them—
Moscovici and Zavalloni (1969) gave participants a seven-point scale, from “strongly disagree” (-3) to “strongly agree” (+3) to report their opinions.
Burnstein and Vinokur (1975) used a ten-point
scale from 1 to 10 representing “the lowest odds of success acceptable” to
try a certain behavior, where 1 means 10\%.  In these and every other survey-based group polarization study I know, the reported opinions on the ordinal scale—with discrete numbered bins—were treated as if they were continuous-valued, or in other words as if they were the internal mental opinion value itself, whatever that might be neurobiologically.

Note how ordinal scales clip all latent opinions with a magnitude greater than the
magnitude of its extreme values. 
lead to false positives because ordinal scales cannot track changes in
extreme opinions beyond the ceiling of the measurement scale—this approach is
ubiquitous in social psychology and survey research, not at all limited to
group polarization.
